%    IPSUR: Introduction to Probability and Statistics Using R
%    Copyright (C)  2018 G. Jay Kerns
%
%    This file is part of IPSUR.
%
%    Permission is granted to copy, distribute and/or modify this document
%    under the terms of the GNU Free Documentation License, Version 1.3
%    or any later version published by the Free Software Foundation;
%    with no Invariant Sections, no Front-Cover Texts, and no Back-Cover Texts.
%    A copy of the license is contained in the LICENSE file in this
%    directory.

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Preface to the Third Edition}

\chapter*{Preface to the Third Edition}

What is new in the Third Edition? Lots of things. In many ways the Second Edition was the testing ground for many experiments. The Third Edition is an abandonment of those experiments and a return to that which works.

\section*{Acknowledgements}

The successes of the Third Edition (if any) will be due in no small part to the successes of the First and Second Editions, so it would be apropos to copy-paste the acknowledgements from the earlier Prefaces here.

I think, though, that the failures of earlier editions have played an important role as well, if not more so. I would like to extend gracious thanks to Mr. P.J.C. Dwarshuis (Hans), Statistician, from The Hague, Netherlands, and Jesus Juan, who, armed with sharp eyes, have pointed out mistakes, misstatements, and places where better discussion is warranted. It is the selfless contributions of people just like these gentlemen which make the hours spent polishing a FREE book all the more worthwhile.

\vfill{}
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures

\vfill{}
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{List of Tables}
\listoftables

